MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-06-27 latex/paper/sections/10-placement.tex
legalization + discussion

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" moves and are akin to physics simulators.

@@

In force directed moves, there is a concept of global placement versus detailed placement, where global placement is driven by forces and detailed placement attempts to legalize the global placement by discretizing it onto the grid of \texttt{Site}s on the device.
There also exists other strategies to mix packing, placement, and legalization in different ways. (CITE THAT ONE PAPER).
In this context, our placer follows a straightforward pack-legalize-place strategy.

If the move reduces the HPWL cost, then the movement is accepted outright.
If the move increases cost, then it can be accepted by chance if the global temperature is high enough to permit the hill-climb.
The temperature decreases with each pass until reaching zero, at which point, the algorithm makes exclusively energy-decreasing moves, effectively reducing into a greedy algorithm. 
The hope is that by this point, the placement has already crystallized into an approximate global optimum and the greediness of the algorithm will help the system settle deeper into the energy well faster. 
We can also switch from random selection to centroid selection when the temperature approaches near zero to speed up the convergence further.

>>>>

In more sophisticated placers, the proposed movement can also be selected by finding the centroid between \texttt{SiteInst}s that share a net with the current object, or even a hybrid of random and centroid selection. 
Moves that actively attempt to predict a better placement are often referred to as "directed" moves in literature. 
Placers that take inspiration from physical interactions like gravity or electrostatics are often referred to as "force directed" placers and are akin to physics simulators.

@@

In force directed placers, there is a concept of global placement followed by detailed placement
Global placement places objects freely on an analog plane and is driven by directed movement or analytical solution.
In global placement, objects do not have to adhere to any hardware constraints and can even overlap with one another.
Detailed placement takes the global placement and attempts to fit it within the constraints of the hardware. 
Detailed placement "legalize" the global placement by discretizing the analog positions of the objects onto the discrete grid of resources of the \texttt{device}.
Overlapping objects are spread out until every \texttt{SiteInst}-\texttt{Cell} is assigned its own unique \texttt{Site}-\texttt{BEL}.

There also exists other strategies to mix packing, placement, and legalization in different ways. (CITE THAT ONE PAPER).
In this context, our placer follows a straightforward pack-legalize-place strategy.

If the move reduces the HPWL cost, then the movement is accepted outright.
If the move increases cost, then it can be accepted by chance if the global temperature is high enough to permit the hill-climb.
The temperature decreases with each pass until reaching zero, at which point, the algorithm makes exclusively energy-decreasing moves, effectively reducing into a greedy algorithm. 
The hope is that by this point, the placement has already crystallized into an approximate global optimum and the greediness of the algorithm will help the system settle deeper into the energy well faster. 
We can also switch from random selection to centroid selection when the temperature approaches near zero to speed up the settlement further.

### 2025-06-27 latex/paper/sections/11-placement_results.tex
legalization + discussion

We will need an HDL design that utilizes a good mixture of LUTs, FFs, BRAMs, and DSPs at scale to demonstrate the robustness and performance of our placers. 
Any DSP subsystem can serve as a good candidate for such a demonstration. 
Here, we will perform placement on a 2048th-order FIR Filter which was conveniently created as coursework for a VLSI course. 
After passing synthesis, the FIR Filter design calls for the following primitive cells:
\begin{itemize}
    \item 1693 \texttt{LUT} (1-6)
    \item 920 \texttt{FDRE} + \texttt{FDSE}
    \item 282 \texttt{CARRY4}
    \item 32 \texttt{RAMB48E1}
    \item 64 \texttt{DSP48E1}
\end{itemize}

@@

Our expected overall utilization for the FIR Filter on this device will be:
\begin{itemize}
    \item LUTs: 3\%
    \item Registers: 1\%
    \item Carry: 2\%
    \item BRAMs: 23\%
    \item DSPs: 29\%
\end{itemize}

@@

The next three placers performed significantly better than the previous two placers.
\texttt{PlacerAnnealRandom} and \texttt{PlacerAnnealHybrid} performed quite similarly with final costs of 346K and 322K respectively.

In between \texttt{AnnealRandom} and \texttt{AnnealHybrid} comes \texttt{PlacerGreedyRandom} with a final cost of 333K.

>>>>

We will need an HDL design that utilizes a good mixture of LUTs, FFs, BRAMs, and DSPs at scale to demonstrate the robustness and performance of our placers. 
Any DSP subsystem can serve as a good candidate for such a demonstration. 
Here, we will perform placement on a 2048th-order FIR Filter which was conveniently created as coursework for a VLSI course. 
After passing synthesis, the FIR Filter design calls for the following primitive cells:

\begin{itemize}
    \item 1693 \texttt{LUT}
    \item 920 \texttt{FF}
    \item 282 \texttt{CARRY4}
    \item 32 \texttt{RAMB48E1}
    \item 64 \texttt{DSP48E1}
\end{itemize}

@@

Our expected overall utilization for the FIR Filter on this device will be:
\begin{itemize}
    \item Lookup Tables: 3\%
    \item Flip Flops: 1\%
    \item Carry: 2\%
    \item BRAMs: 23\%
    \item DSPs: 29\%
\end{itemize}

@@

The next three placers performed significantly better than the previous two placers.
Our vanilla SA placer, \texttt{PlacerAnnealRandom} is represented by the green curve.
We can observe that the cost decreases more steadily than the previous two placers, with noticeable spikes in total system cost due to hill-climbing, until settling at the 346K mark as the global temperature cools to zero.

Next is \texttt{PlacerGreedyRandom} which is represented by the purple curve.
We can observe a much steeper initial drop in system cost than the vanilla SA placer due to greedy movement acceptance.
Interestingly, this placer also settled at a lower final system cost of 333K, which is a slight 4\% improvement over vanilla SA.
We originally expected the greediness to be a detriment to finding the global minimum, but due to the inherent randomness of the algorithm, these expectations are never certain 
Broader trends will require many more test trials to observe reliably.

The best performing placer was \texttt{PlacerAnnealHybrid}, represented by the blue curve.
We can observe that \texttt{AnnealHybrid} settles at a similar rate with the vanilla \texttt{AnnealRandom}, but eventually settles at a lower minimum at 322K, outperforming even \texttt{GreedyRandom}.
